Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-30 Report/Sections/data.tex
Update Report

Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, which can be helpful for conservation efforts. However, studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the energetic requirements of killer whales. This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.

We begin by describing the data collection process used by ecologists. We then detail the procedure used to build several candidate generative models from this data. Next, we perform a simulation study by selecting a preferred model, simulating data from it, and fit the candidate models to the simulated data to compare their performance in terms of parameter estimates and decoding accuracy. Finally, we fit each candidate model to the original data and analyze the results.

@@

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods were ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. Figure \ref{fig:data} displays the dive profile and accelerometer data.

Each dive is treated as one curve, and the collection of dives make up the coarse-scale process. As such, the coarse-scale observations $Y$ were a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$. In addition, the fine-scale processes $Y^*$ were the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive types $X^*$.

@@

The coarse-scale level exhibited no significant amount of auto-correlation (See section S-2.1) and no obvious intricate structure. As such, we selected a simple HMM to model this data. The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)},$$
$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$

The fine-scale acceleration data exhibited significant sinusoidal behavior, thus the fine-scale observations $y^*$ were transformed to $z^*$ using the DFT summary statistics of a two-second sliding window. A two-second window was selected heuristically since it appeared to be the smallest window size that could capture the sinusoidal behavior. Because the data was collected at the rate of 50 Hz, each two-second window contained a total of 100 raw acceleration readings. The acceleration data here was collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. As a result, the transformation from the raw accelerometer data $\mathbf{y}^*$ to the observation sequence $z^*$ was calculated as follows:
%
$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
%
Each observation was therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar. We calculated $z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.

Even after accounting for fine-scale structure in the raw acceleration data, there was still strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ (section S-2.1), thus a CarHMM was used to model $\mathbf{Z}^{*(1)}_{t,t^*}$. In particular, we assumed $\mathbf{Z}^{*(1)}_{t,t^*}$ to be Normally distributed with unknown parameters $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.
%
%$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)},$$
%$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right].$$
%

While $Z^{*(2)}$ also exhibits some auto-correlation, the relationship is less strong than that of $\mathbf{Z}^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear. Therefore, a simple HMM was used to model $Z^{*(2)}$ and the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean $\mu_2^{*(i)}$ and variance $\sigma_2^{*(i)}$.
%
%$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
%$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%
The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ shared a subdive state $X^*_{t,t^*}$, but were assumed to be independent of one another when conditioned on $X_{t,t^*}$. In addition, the distribution of $Z^*_{t,t^*}$ is assumed to depend only upon the subdive type and the previous observation $Z^*_{t,t^*-1}$, not the dive type $X_t$.

Finally, to select the number of dive types and subdive types, information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}. Instead, we selected $N = 2$ dive types and $N^* = 3$ subdive behaviors heuristically. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is a hierarchical hidden Markov model with a traditional HMM at the coarse scale. On the fine scale, $y^*$ was transformed to $z^*$, $z^{*(1)}$ was modeled with a CarHMM, and $z^{*(2)}$ was modeled with an HMM. We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed here. The parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
    %
    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \Theta^{*(i)} =  \{\{\mu_1^{*(i)},\sigma_1^{*(i)},\phi_1^{*(i)}\},\{\mu_2^{*(i)},\sigma_2^{*(i)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
\end{gather*}
%
The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model.

In addition to the preferred CarHHMM-DFT, we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $\mathbf{Z}^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations \textit{without} using the DFT, and models $Z^{*(1)}$ using a CarHMM. The Fourier sums $\mathbf{Z}^{*(2)}$ are omitted as observations in this model.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $\mathbf{Z}^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
All models have the same emission distributions as the CarHHMM-DFT, when applicable. Each of the three models leaves out one important aspect of the full CarHHMM-DFT. The HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations, and the CarHMM-DFT lacks a hierarchical structure by only modelling the fine-scale behavior of the whale.

>>>>

Understanding animal behaviour can be important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}. NRKW prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKW for predation \citep{Williams:2009,Noren:2011}. It is therefore of interest to determine the energetic expenditure of NRKW to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.

Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the true energetic requirements of NRKW.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods were ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contains a total of 267 dives. Figure \ref{fig:data} displays the dive profile and accelerometer data.

Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. As such, the coarse-scale observations $Y$ are a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$. In addition, the fine-scale processes $Y^*$ are the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $X^*$.

@@

A lag plot was constructed to determine where auto-correlation should be incorporated into the model (see Section S-2.1). Prior to fitting the model, the coarse-scale level exhibited no significant amount of auto-correlation and no obvious intricate structure. As such, we selected a simple HMM to model this data. The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:

$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$

The fine-scale acceleration data exhibited significant sinusoidal behavior; thus the fine-scale observations $Y^*$ were transformed to $Z^*$ using the DFT summary statistics of a two-second sliding window. A two-second window was selected heuristically since it appeared to be the smallest window size that could capture the. ``wiggles" in their entirety. Because the data was collected at the rate of 50 Hz, each two-second window contained a total of 100 raw acceleration readings. The acceleration data here was collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. As a result, the transformation from the raw accelerometer data $\mathbf{Y}^*$ to the observation sequence $Z^*$ is calculated as follows:
%
\begin{equation*}
    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right) \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2
\end{equation*}
%
Each observation is therefore made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar. We calculated $Z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.

Even after accounting for fine-scale structure in the raw acceleration data, there was still strong auto-correlation within $Z^{*(1)}_{t,t^*}$ prior to fitting the model (section S-2.1). Therefore, a CarHMM was used to model $Z^{*(1)}_{t,t^*}$. In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ to be Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
%
%$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)},$$
%$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right].$$
%

While $Z^{*(2)}$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $Z^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear. In addition, much of the auto-correlation that \textit{is} evident from the lag plot may be explained by persistence in the subdive type $X^*_{t,t^*}$. Therefore, a simple HMM was used to model $Z^{*(2)}$. In addition, given that $X^*_{t,t^*} = i^*$ the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
%
%$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
%$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%

We do not use information criteria to select the number of dive types and subdive types since it tends to overestimate the number of states in biological processes \citep{Pohle:2017}. In the particular case of well-separated hidden states, a lag plot will reveal $N$ circular patterns for each dive feature and $N^*$ circular patterns for each subdive feature. This is unfortunately not the case for the killer whale data, but we still consult the lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviors. The absence of a more principled method to select the number of hidden states highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is a hierarchical hidden Markov model with a traditional HMM at the coarse scale. On the fine scale, $Y^*$ is transformed to $Z^*$, $Z^{*(1)}$ is modeled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM. In addition, we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity. This implies that dive types 1 and 2 share subdive states. We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed here. The parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
    %
    \Theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\{\mu_1^{*(\cdot,i^*)},\sigma_1^{*(\cdot,i^*)},\phi_1^{*(\cdot,i^*)}\},\{\mu_2^{*(\cdot,i^*)},\sigma_2^{*(\cdot,i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
\end{gather*}
%
Remember that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model for the CarHHMM-DFT. Additionally, we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $Z^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ using a CarHMM. However, the Fourier sums $Z^{*(2)}$ are omitted as observations in this model. The Fourier sums $Z^{*(2)}$ describe the ``wiggliness" of the accelerometer data, while $Z^{*(1)}$ is simply the average acceleration over a particular window.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $\mathbf{Z}^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
All models have the same emission distributions as the CarHHMM-DFT, when applicable. Each of the three models leaves out one important aspect of the full CarHHMM-DFT. The HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations, and the CarHMM-DFT lacks a hierarchical structure by only modelling the fine-scale behavior of the whale.

### 2020-07-30 Report/Sections/discussion.tex
Update Report

Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state. However, many real-world processes require more complex models. Here, we presented a flexible framework that can account for complicated dependence structures within time-series data. Our framework is a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers.

The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM \citep{Lawler:2019}. CarHMMs also maintain the structure needed to evaluate the likelihood using the forward algorithm. In our Normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state. Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within an observation sequence.

Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals. A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citet{Michelot:2019}. Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result are not easily incorporated into the HHMM structure. However, we show in Theorem 1 (see appendix) that the state-switching OU process is equivalent to an CarHMM under certain conditions. This implies that the parameter estimates from the CarHMM are interpretable in the context of a stochastic differential equation.

For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs. In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$. \citet{Adam:2019} utilize the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California. However, they run into clear violations of conditional independence and apparent periodic behavior when plotting the auto-correlation of acceleration as a function of lag, implying the need to account for complicated dependence structures within fine-scale processes with high sampling frequencies.

%For processes such as these with very high sampling frequencies and/or intricate fine-scale structure, 
To deal with these issues, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood. For example, if the sampling rate of the fine-scale process is very high and clearly period, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.

In particular, we use $Z^{*(1)}$ and $Z^{*(2)}$ based on the DFT as defined previously. These summary statistics prove to be vital when decoding the behavior of a killer whale. This is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and the case study, where the CarHHMM without DFT visually fails to pick up fluking behavior in the killer whale.

>>>>

Current FDA literature largely address sequences of curves either as multilevel models with random effects and no temporal dependence \cite{chen:2012,di:2009} or with functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \cite{Kokoszka:2018}. However, many real-world curve sequences exhibit temporal dependence and a discrete number of ``types". One obvious example is the case study examined here, where a sequence of marine mammal dives serve a number of discrete of purposes such as traveling, socializing, resting, and foraging \citep{Tennessen:2019a}. One other example can be found when measuring accelerometer data in school children, where bouts of activity can be associated with running, walking, etc. \citep{Morris:2007}.

To deal with both temporal dependence between curves and a discrete number of curve types, we suggest modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation within the data. Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within an sequence of curves. One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use another within-curve parametric model as emissions. The parameters of this fine-scale model in turn will depend upon the curve type, and together these two levels make up a hierarchical model which can simultaneously account for processes taking place at different time-scales.

%Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state. However, many real-world processes require more complex models. Here, we presented a flexible framework that can account for complicated dependence structures within time-series data. Our framework is a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers.

%The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM \citep{Lawler:2019}. CarHMMs also maintain the structure needed to evaluate the likelihood using the forward algorithm. In our Normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state.

Several possible choices exist for the fine-scale model, including yet another HMM. \citet{Barajas:2017} introduce this model in the context of statistical ecology as an HHMM, and \citet{Adam:2019} use it to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California. However, traditional HMMs prove insufficient at very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and apparent periodic behavior in the acceleration data. This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.

%For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs. In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.

%For processes such as these with very high sampling frequencies and/or intricate fine-scale structure,

%To deal with these issues, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood. For example, if the sampling rate of the fine-scale process is very high and clearly period, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.

To deal with these problematic fine-scale processes, we suggest using a moving-window transformation of the fine-scale observations. For periodic behaviour in particular we use a DFT-based transformation. The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and our case study, where the CarHHMM without DFT visually fails to pick up a killer whale's clear fluking behaviour.

We also suggest using a CarHMM instead of a traditional HMM for the fine-scale process when expert opinion and careful observation (possibly using a log plot) calls for it. Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals. A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citet{Michelot:2019}. Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result are not easily incorporated into the HHMM structure. However, we show in Theorem 1 (see appendix) that the state-switching OU process is equivalent to an CarHMM under certain conditions. This implies that the parameter estimates from the CarHMM are interpretable in the context of a stochastic differential equation.

### 2020-07-30 Report/Sections/intro.tex
Update Report

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model this behaviour by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, for the fine-scale behaviour, we use summary information that is biologically meaninful.   Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

@@

We also borrow techniques from animal movement to model fine-scale behaviour, beyond simply using an HMM.  
Under the traditional HMM model,  observations are independent given  the underlying hidden state process.
However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
Several solutions have been proposed, such as  the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
The CarHMM  explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.

An additional complication is that, at the shortest time scales,  observations often exhibit complicated  structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}. 
%One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
Our dive data exhibit  fine-scale periodic behaviour, so we transform the data via
 the signal processing technique of the Fourier transform. 
\cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM  in the context of describing daily behavioral cycles of marine mammals.
In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data. 
Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.

This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models. 
In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
We then perform a simulation study to compare the performance of each existing model with ours in a controlled setting. 
Finally, we apply our method to the dive data of Figure 6.

In the following section, we describe HMMs, CarHMMs and HHMMs.  We show how these models can be combined in order to analyze complex data. We also discuss transforming data.  In Section 3, we study various models for complex ...  Section 4 contains data analysis and Section 5 discussion.

>>>>

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model this behaviour by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, we use a biologically meaningful transformation on the fine-scale behaviour to account for the ``wiggliness'' in some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state. % Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

@@

We also borrow techniques from animal movement to model within-dive behaviour, beyond simply using an HMM.  
Under the traditional HMM model,  observations are independent given  the underlying hidden state process.
However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
Several solutions have been proposed, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
The CarHMM explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.

An additional complication seen in the ecology literature is that, at the shortest time scales, observations often exhibit complicated structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs. One example in addition to the case study from this work is the swimming patterns of horn sharks off the coast of Southern California \citep{Adam:2019}, both of which exhibit fine-scale periodic behavior. 
%One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
%Our dive data exhibit fine-scale periodic behaviour, so we transform the data via the Fourier transform. 
\cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM in the context of describing daily behavioral cycles of marine mammals.
In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior using accelerometer data \citep{Fehlmann:2017,Shorter:2017}.
Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.

%This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
%We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models. 
%In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
%We then perform a simulation study to compare the performance of each existing model with ours in a controlled setting.

In the following section, we describe HMMs, generalize them to CarHMMs, introduce hierarchical structure with HHMMs, and discuss methods to transform intricate fine-scale data. We also show how these models can be combined in order to analyze complex data. In Section 3, we construct several candidate models based on the data from Figure \ref{fig:data} and Section 4 details a simulation study based on these candidate models. In Section 5 we fit each candidate model to the original data and analyze the results, and Section 6 contains our discussion of the results.
